---
output:
  md_document:
    variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-figs/",
  cache.path = "README-cache/"
)
```

<style>
.nav{
    border:1px solid #ccc;
    border-width:1px 0;
    list-style:none;
    margin:0;
    padding:0;
    text-align:center;
}
.nav li{
    display:inline-block;
}
.nav a{
    display:inline-block;
    padding:5px;
}
</style>

<ul class="nav">
  <li><a href="#install">Install</a></li>
  <li><a href="#documentation">Documentation</a></li>
  <li><a href="#example">Example</a></li>
  <li><a href="#citation">Citation</a></li>
  <li><a href="#license">License</a></li>
</ul>

This package has the data for the University of Washington Fish507 Applied Time Series Analysis [course](https://nwfsc-timeseries.github.io/atsa) and [lab book](https://nwfsc-timeseries.github.io/atsa-labs).


### INSTALL {#install}

To install:
```
library(devtools)
devtools::install_github("nwfsc-timeseries/atsalibrary")
```

Windows users:

You may need to turn off warning-error-conversion, because since the R 4.0 update, the tiniest warning stops installation of GitHub hosted packages.
```
Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
```

### EXAMPLE {#example}

Type `?` and the dataset name for the help file for each dataset.

```{r}
data(package="atsalibrary")
```

### DOCUMENTATION {#documentation}

- [ATSA lab book](https://nwfsc-timeseries.github.io/atsa-labs/) - The atsalibrary is used in our Applied Time Series Analysis book developed from the labs in our course.
- [ATSA course website](https://nwfsc-timeseries.github.io/atsa/) - We have lectures and all material from our course on our course website.

### CITATION {#citation}

Holmes, E.E., M.D. Scheuerell, and E.J. Ward. 'atsalibrary': Applied Time Series Analysis datasets. Version `r packageVersion("atsalibrary")`.

### LICENSE {#license}

atsalibrary was developed by US federal government employees as part of their official duties. As such, it is not subject to copyright protection and is considered “public domain” (see 17 USC § 105). Public domain software can be used by anyone for any purpose, and cannot be released under a copyright license.